A comparison of design-based and model-based approaches for spatial data.

### 2021-09-03 inst/SpatialDVM_Manuscript/SpatialDVM_Manuscript.Rmd
incorporated revisions from Lisa.

@@

The design-based approach assumes the data are fixed. Randomness is incorporated in the selection of population units according to a sampling design. A sampling design assigns a positive probability of inclusion in the sample (inclusion probability) to each population unit. Some examples of commonly used sampling designs include independent random sampling (IRS), stratified random sampling, and cluster sampling. The goal is to use the sampling design and the sampled data to estimate population parameters like means and totals. These population parameters are typically assumed to be fixed but unknown.

@@

The model-based approach assumes the data are a random realization of a data-generating process. Randomness is often incorporated through distributional assumptions on this process. Instead of estimating fixed but unknown parameters (as in the design-based approach), the goal of model-based inference in the spatial context is often \emph{prediction} of an unknown quantity. For example, suppose the realized mean of all population units is the quantity of interest. Instead of \emph{estimating} a fixed unknown mean, we are \emph{predicting} the value of the mean, a random variable. We know that if we sampled all population units, we would have an exact prediction for the mean of our one realized process, without any uncertainty. But we are typically not interested in the true, unknown mean of the underlying process.

@@

```{r fig1, out.width = "100%", fig.cap = "A comparison of sampling under the design-based and model-based frameworks. In the top row, we have one fixed population, and two random samples. In the bottom row, we have two realizations of the same spatial process sampled at the same locations."}
set.seed(08032021)

@@

@@

Spatially balanced samples are useful because they tend to yield estimates that have lower variance than estimates constructed from sampling designs lacking spatial balance [@stevens2004spatially; @barabesi2011sampling; @grafstrom2013well; @robertson2013bas; @wang2013design; @benedetti2017spatiallyreview]. To quantify spatial balance, @stevens2004spatially proposed loss functions based on Voroni polygons. The first spatially balanced sampling algorithm that saw widespread use was the Generalized Random Tessellation Stratified [@stevens2004spatially]. Since GRTS was developed, several other spatially balanced sampling algorithms have emerged, including the Local Pivotal Method [@grafstrom2012spatially; @grafstrom2018spatially], Spatially Correlated Poisson Sampling [@grafstrom2012spatiallypoisson], Balanced Acceptance Sampling [@robertson2013bas], Within-Sample-Distance [@benedetti2017spatially], and Halton Iterative Partitioning [@robertson2018halton]. We focus on the Generalized Random Tessellation Stratified (GRTS) algorithm to select spatially balanced sampling because it has several attractive properties detailed by @stevens2004spatially and @dumelle2021spsurvey.

@@

The GRTS algorithm is used to sample from finite and infinite populations and works by utilizing a mapping between two-dimensional and one-dimensional space. The population units in two-dimensional space are divided into cells using a hierarchical index. Population units are then mapped to a one-dimensional line via the hierarchical indexing. The line length of each population unit equals its inclusion probability. A systematic sample is conducted on the line and these samples are linked to a population unit in two-dimensional space, which results in the desired sample. @stevens2004spatially provide and @dumelle2021spsurvey provide further details.

@@

Finite Population Block Kriging (FPBK) is a model-based approach that expands the geostatistical Kriging framework to the finite population setting [@verhoef2008spatial]. Instead of basing inference off of a specific sampling design, we assume the data are generated by a spatial process. @verhoef2008spatial gives details on the theory of FPBK, but some of the basic principles are summarized below. Let $\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}$ be a response variable that can be measured at the $N$ population units and is represented as an $N \times 1$ vector. Suppose we want to predict some linear function of the response variable, $f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}$, where $\mathbf{b}$ is a $1 \times N$ vector of weights. For example, if we want to predict the population total across all population units, then we would use a vector of 1's for the weights.

@@

Typically, however, we only have a sample of the $N$ population units. Denoting quantities that are part of the sampled population units with a subscript \emph{s} and quantities that are part of the unsampled population units with a subscript \emph{u},

@@

We also typically assume that there is spatial correlation in $\bm{\delta}$, which can be modeled using a covariance function. It is common to assume the covariance function is second-order stationary and isotropic [@cressie1993statistics], and that the spatial covariance decreases as the separation between population units increases. Many spatial covariance functions exist, but the primary function we use throughout the simulations and applications in this manuscript is the exponential covariance function: the $i,j^{th}$ entry for $\cov(\bm{\delta})$ is
\mbox{}
\begin{align}\label{equation:expcov}
\cov(\delta_i, \delta_j) = \theta_1\exp(-3h_{i,j}/\theta_2) + \theta_3\mathbbm{1}\{\mathbf{h}_{i,j} = 0\}, 
\end{align}
where $h_{i,j}$ is the distance between population units $i$ and $j$, and $\bm{\theta}$ is a vector of spatial covariance parameters of the partial sill $\theta_1$, the range $\theta_2$, and the nugget $\theta_3$, and $\mathbbm{1}$ is an indicator function. However, any spatial covariance function could be used in the place of the exponential, including functions that allow for non-stationarity or anisotropy [@chiles1999geostatistics, pp. 80-93].

@@

We note that we only use FPBK in this paper in order to focus more on comparing the design-based and model-based approaches. However, k-nearest-neighbors [@fix1951discriminatory; @ver2013comparison], random forest [@breiman2001random], Bayesian models [@chan2020bayesian], among others, can also be used to obtain predictions for a mean or total from spatially correlated responses in a finite population setting.

@@

* correlation type: dependent errors or independent errors
* error type:
    * normal: mean 0, variance 2
    * lognormal: log scale mean 0, log scale variance 2 (total variance 47)
    * lognormalbig: log scale mean 0, log scale variance 4 (total variance 2,926)
* sample sizes: $n = 10, 50, 150; N = 900$
* layout: gridded vs random population locations

@@

In each simulation, a GRTS sample and an IRS sample were selected. Then for the GRTS sample, the design-based approach using the local neighborhood variance (Design GRTS) and a model-based approach were applied (Model GRTS). Then for the IRS sample, the design-based approach using the simple random sample variance (Design IRS) and a model-based approach were applied (Model IRS).

@@

2. We will now focus in a bit more on comparing Design-GRTS to Model-GRTS, the two best approaches for any reasonable sample size. In the independent error settings, the two approaches perform very similarly, so those results are omitted in the following graph. In the dependent error settings, using rmspe as the performance criterion, Model-GRTS outperforms Design-GRTS in 12 of the 18 settings, the two approaches perform very similarly in 3 settings, and Design-GRTS outperforms Model-GRTS in 3 settings.

@@

The Environmental Protection Agency (EPA), states, and tribes periodically conduct National Aquatic Research Surveys (NARS) in the United States to assess the water quality of various bodies of water. We will use the 2012 National Lakes Assessment (NLA), which measures various aspects of lake health and quality in lakes in the contiguous United States, to obtain an interval for mean mercury concentration. Although all lakes in the survey were measured in 2012, there may not always be enough time or money to do so. Therefore, we will explore whether or not we can still obtain a relatively precise estimate for the realized mean mercury concentration if we only take a sample of 100 of the 986 lakes.

@@

Figure \ref{fig:figdata} shows that mercury concentration is right-skewed, with most lakes having a low value of mercury concentration but a few having a much higher concentration. Mercury concentration exhibits some spatial correlation, with high mercury concentrations in lakes in the northeast and north central United States. Because we are considering these lakes to be our entire population, we know that the realized mean mercury concentration is 103.03 ng / g.

@@

```{r apptab}
library(knitr)
names(res_df) <- c("Approach", "Realized Mean", "Estimate", "SE", "95% LB", "95% UB")
kable(res_df, digits = 1, caption = "\\label{tab:appliedtab} Table XXX. Application of design-based and model-based approaches to the NLA data set on mercury concentration.")
```

>>>>

Lisa M.: Add paragraph describing contribution of manuscript.

@@

The design-based approach assumes the population is fixed. Randomness is incorporated in the selection of population units according to a sampling design. A sampling design assigns a positive probability of inclusion in the sample (inclusion probability) to each population unit. Some examples of commonly used sampling designs include independent random sampling (IRS), stratified random sampling, and cluster sampling. The goal is to use the sampling design and the sampled data to estimate population parameters like means and totals. These population parameters are traditionally assumed to be fixed but unknown.

@@

The model-based approach assumes the data are a random realization of a data-generating process. Randomness is often incorporated through distributional assumptions on this process. Instead of estimating fixed but unknown parameters (as in the design-based approach), the goal of model-based inference in the spatial context is often \emph{prediction} of an unknown quantity. For example, suppose the realized mean of all population units is the quantity of interest. Instead of \emph{estimating} a fixed unknown mean, we are \emph{predicting} the value of the mean, a random variable. We know that if we sampled all population units, we would have an exact prediction for the mean of our one realized process, without any uncertainty. But we are often not interested in the true, unknown mean of the underlying process.

@@

```{r fig1, out.width = "100%", fig.cap = "A comparison of sampling under the design-based and model-based frameworks. Points circled are those that are sampled. In the top row, we have one fixed population, and two random samples. In the bottom row, we have two realizations of the same spatial process sampled at the same locations."}
set.seed(08032021)

@@

Lisa M.: Need a more precise definition of "miniature" in this context, and need an example.

@@

Spatially balanced samples are useful because they tend to yield estimates that have lower variance than estimates constructed from sampling designs lacking spatial balance [@stevens2004spatially; @barabesi2011sampling; @grafstrom2013well; @robertson2013bas; @wang2013design; @benedetti2017spatiallyreview]. To quantify spatial balance, @stevens2004spatially proposed loss functions based on Voroni polygons. The first spatially balanced sampling algorithm that saw widespread use was the Generalized Random Tessellation Stratified [@stevens2004spatially]. Since GRTS was developed, several other spatially balanced sampling algorithms have emerged, including the Local Pivotal Method [@grafstrom2012spatially; @grafstrom2018spatially], Spatially Correlated Poisson Sampling [@grafstrom2012spatiallypoisson], Balanced Acceptance Sampling [@robertson2013bas], Within-Sample-Distance [@benedetti2017spatially], and Halton Iterative Partitioning [@robertson2018halton]. We focus on the Generalized Random Tessellation Stratified (GRTS) algorithm to select spatially balanced sampling because it has several attractive properties, including Lisa M.: List major attractive properties, and detailed by @stevens2004spatially and @dumelle2021spsurvey.

@@

The GRTS algorithm is used to sample from finite and infinite populations and works by utilizing a mapping between two-dimensional and one-dimensional space. The population units in two-dimensional space are divided into cells using a hierarchical index. Population units are then mapped to a one-dimensional line via the hierarchical indexing. The line length of each population unit equals its inclusion probability. A systematic sample is conducted on the line and these samples are linked to a population unit in two-dimensional space, which results in the desired sample. @stevens2004spatially and @dumelle2021spsurvey provide further details.

@@

Finite Population Block Kriging (FPBK) is a model-based approach that expands the geostatistical Kriging framework to the finite population setting [@verhoef2008spatial]. Instead of basing inference off of a specific sampling design, we assume the data are generated by a spatial process. @verhoef2008spatial gives details on the theory of FPBK, but some of the basic principles are summarized below. Let $\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}$ be a response variable that can be measured at the $N$ population units and is represented as an $N \times 1$ vector. Suppose we want to predict some linear function of the response variable, $f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}$, where $\mathbf{b}^\prime$ is a $1 \times N$ vector of weights. For example, if we want to predict the population total across all population units, then we would use a vector of 1's for the weights.

@@

However, we often only have a sample of the $N$ population units. Denoting quantities that are part of the sampled population units with a subscript \emph{s} and quantities that are part of the unsampled population units with a subscript \emph{u},

@@

We also assume that there is spatial correlation in $\bm{\delta}$, which can be modeled using a covariance function. It is common to assume the covariance function is second-order stationary and isotropic [@cressie1993statistics], and that the spatial covariance decreases as the separation between population units increases. Many spatial covariance functions exist, but the primary function we use throughout the simulations and applications in this manuscript is the exponential covariance function: the $i,j^{th}$ entry for $\cov(\bm{\delta})$ is
\mbox{}
\begin{align}\label{equation:expcov}
\cov(\delta_i, \delta_j) = \theta_1\exp(-3h_{i,j}/\theta_2) + \theta_3\mathbbm{1}\{\mathbf{h}_{i,j} = 0\}, 
\end{align}
where $h_{i,j}$ is the distance between population units $i$ and $j$, and $\bm{\theta}$ is a vector of spatial covariance parameters of the partial sill $\theta_1$, the range $\theta_2$, and the nugget $\theta_3$; and, $\mathbbm{1}$ is equal to 1 when distance $h_i,j$ is equal to 0, and equal to 0 otherwise. However, any spatial covariance function could be used in the place of the exponential, including functions that allow for non-stationarity or anisotropy [@chiles1999geostatistics, pp. 80-93].

Lisa M. : Include formulas. Perhaps, but, these are very heavy in notation and matrix algebra. We might consider, however, adding the formulas to an Appendix.

@@

We note that we only use FPBK in this paper in order to focus more on comparing the design-based and model-based approaches. However, k-nearest-neighbors [@fix1951discriminatory; @ver2013comparison], random forest [@breiman2001random], Bayesian models [@chan2020bayesian], among others, can also be used to obtain predictions for a mean or total from spatially correlated responses in a finite population setting. We choose to use FPBK because it is faster than a Bayesian approach and random forest and because @ver2013comparison showed that the method outperforms k-nearest-neighbors in many scenarios.

@@

* correlation type: dependent errors or independent errors
* error type:
    * normal: mean 0, variance 2
    * lognormal: log scale mean 0, log scale variance 2 (total variance 47)
    * lognormalbig: log scale mean 0, log scale variance 4 (total variance 2,926)
* sample sizes: $n = 10, 50, 150; N = 900$
* layout: gridded vs random uniform population locations confined to a 1 x 1 unit square

@@

Lisa M. Notes: adding an intermediate level of spatial dependency?
Transfer simulation scenarios to a table
Explain what the effective range is
Fill in the details of what exactly each approach means (perhaps a table would be a good way to do this)
Reorder the sims in the first figure by some criterion.
Think about what would be a "reasonable sample size" instead of 10.
Define medae
If the data has a large right-skew, wouldn't one consider a transformation before the analysis? We should address this by stating that the BLUP for the log response does not mean that e^logBLUP is the BLUP for the response on the original scale.

In each simulated data set, a GRTS sample and an IRS sample were selected. Then for the GRTS sample, the design-based approach using the local neighborhood variance (Design GRTS) and a model-based approach were applied (Model GRTS). Then for the IRS sample, the design-based approach using the simple random sample variance (Design IRS) and a model-based approach were applied (Model IRS).

@@

2. We will now focus on comparing Design-GRTS to Model-GRTS, the two best approaches for any reasonable sample size. In the independent error settings, the two approaches perform very similarly, so those results are omitted in the following graph. In the dependent error settings, using rmspe as the performance criterion, Model-GRTS outperforms Design-GRTS in 12 of the 18 settings, the two approaches perform very similarly in 3 settings, and Design-GRTS outperforms Model-GRTS in 3 settings.

@@

The Environmental Protection Agency (EPA), states, and tribes periodically conduct National Aquatic Research Surveys (NARS) in the United States to assess the water quality of various bodies of water. We will use the 2012 National Lakes Assessment (NLA), which measures various aspects of lake health and quality in lakes in the contiguous United States, to obtain an interval for mean mercury concentration. Although all lakes in the survey were measured in 2012, there may not always be enough time or money to do so. Therefore, we will explore whether or not we can still obtain an adequately precise estimate for the realized mean mercury concentration if we only take a sample of 100 of the 986 lakes.

@@

Figure \ref{fig:figdata} shows that mercury concentration is right-skewed, with most lakes having a low value of mercury concentration but a few having a much higher concentration. Mercury concentration exhibits some spatial correlation, with high mercury concentrations in lakes in the northeast and north central United States. Because we are considering these lakes to be our entire population, we know that the realized mean mercury concentration is 103.2 ng / g.

@@

```{r apptab}
library(knitr)
res_df_drop <- res_df[ ,-2]
names(res_df_drop) <- c("Approach", "Estimate", "SE", "95% LB", "95% UB")
kable(res_df_drop, digits = 1, caption = "\\label{tab:appliedtab} Application of design-based and model-based approaches to the NLA data set on mercury concentration. The true mean concentration is 103.2 103.2 ng / g")
```
